\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}

\geometry{
  top=1.2in,
  bottom=1.5in,
  left=1.1in,
  right=1.1in,
  headheight=10pt,
  headsep=10pt,
  footskip=50pt
}

\usepackage{graphicx}
\usepackage{lipsum} 
\usepackage{xcolor} 
\usepackage[hidelinks]{hyperref}
\usepackage{multirow}

\usepackage{caption}
\usepackage{float}
\usepackage{mathtools}

\usepackage{titlesec}
\titleformat{\section}       % Command to format sections
  {\normalfont\bfseries\fontsize{9}{10}\selectfont}          % Font size and style for section titles
  {\thesection}              % Section number format
  {0.5em}                      % Spacing between number and title
  {}                         % Code before the title

\titleformat{\subsection}  % Command to format subsections
  {\normalfont\bfseries\fontsize{9}{10}\selectfont}  % Font size and style for subsection titles
  {\thesubsection}          % Subsection number format
  {0.5em}                   % Spacing between number and title
  {}                        % Code before the title

% Tags for the mat expressions, use [] instead
\newtagform{brackets}{[}{]}
\usetagform{brackets}

\title

\begin{document}

\noindent
\textbf{Psychological Factors and COVID-19 Test Outcomes: Associations and Differences}\\
\vspace{0.1em} 
\noindent\hrulefill 
\vspace{0.8em} 

\begin{center}
    \textbf{Eurico Martins, Gutelvam Fernandes, Hugo Alves}\\
    \vspace{0.5em}
    Master in Applied Artificial Intelligence;\\
		EST - Instituto Politécnico do Cávado e do Ave, Portugal.\\
    \vspace{0.5em}
    \href{mailto:a8794@alunos.ipca.pt}{a8794@alunos.ipca.pt}, \href{mailto:a33791@alunos.ipca.pt}{a33791@alunos.ipca.pt}, \href{mailto:a30783@alunos.ipca.pt}{a30783@alunos.ipca.pt}
\end{center}

\vspace{3em}
\noindent {\bfseries \fontsize{9}{11}\selectfont ABSTRACT}
\vspace{1em}\newline
This study investigates the relationship between psychological factors—stress, anxiety, depression, and optimism—and COVID-19 test outcomes. Correlation analyses identified significant associations among the psychological variables, though their correlations with testing positive for COVID-19 were weak. Stress, anxiety, and depression exhibited minimal negative associations, while optimism demonstrated a positive correlation with positive test results. Normality tests revealed deviations from normality, and Mann-Whitney U tests indicated significant group differences in stress, depression, and optimism, while anxiety showed no statistically significant difference. These findings highlight nuanced links between psychological health and COVID-19 testing outcomes.

\vspace{1em}
\textbf{Keywords}: Psychological factors, stress, anxiety, depression, COVID-19

\vspace{2em}
\section{INTRODUCTION}
\vspace{0.5em}
The COVID-19 pandemic has had a profound impact on both physical and mental health worldwide.
While the physical health consequences are well-documented, the psychological effects have also been significant, with heightened levels of depression, anxiety, and stress reported across the globe.
Social isolation, economic uncertainty, and ongoing health fears have all contributed to a marked increase in mental health challenges, exacerbating existing vulnerabilities.
The pandemic disrupted daily routines, created widespread financial and social stress, and fostered an environment of uncertainty, leading to substantial deterioration in psychological well-being for many individuals.
\vspace{0.5em}\newline
Depression emerged as one of the most prevalent psychological effects of the pandemic.
Social isolation, coupled with uncertainty surrounding the virus, intensified feelings of helplessness and loneliness. Economic challenges, such as job losses and financial insecurity, further contributed to the surge in depressive symptoms.
The necessary social distancing measures, though essential for preventing virus spread,
left many without the support systems that might have helped mitigate feelings of isolation, leading to a global rise in depression during the pandemic.
\vspace{0.5em}\newline
Anxiety also became a major issue, driven by the constant flow of information about the virus and the fear of its unpredictable impact on health.
The concern about personal and familial well-being, along with economic instability, led to heightened anxiety levels.
Unlike transient anxiety, the persistent nature of the pandemic resulted in chronic distress, affecting individuals' ability to cope with daily life and function normally.
These prolonged periods of anxiety added to the overall mental health burden of the crisis.
\vspace{0.5em}\newline
Dispositional optimism, or the general expectation that good things will happen, emerged as a key protective factor against the mental health challenges of the pandemic.
Optimistic individuals tend to cope with adversity in more adaptive ways, maintaining a sense of hope and agency during difficult times.
Research has shown that optimism is associated with better psychological outcomes, including lower stress levels and greater resilience. During the pandemic,
optimism allowed individuals to manage uncertainty more effectively, promoting healthier behaviors and improving overall well-being.
\vspace{0.5em}\newline
Stress, an inevitable response to perceived challenges, has been another significant consequence of the pandemic.
The disruption of daily life, fear of illness, and uncertainty about the future combined to create a chronic state of stress for many. Prolonged stress has been linked to various physical and mental health issues,
including cardiovascular problems and weakened immune systems, further exacerbating the psychological toll of the pandemic.
The persistent nature of stress during this period has contributed to a rise in stress-related disorders globally.
\vspace{0.5em}\newline
The intersection of COVID-19 with mental health issues like depression, anxiety, and stress highlights the need for psychological resilience.
While the pandemic has undeniably worsened global mental health, optimism has proven to be a key factor in mitigating its effects. Further research into 
how optimism influences mental health could offer valuable strategies to strengthen resilience and better equip individuals to cope with the ongoing challenges posed by global crises.


\vspace{2em}
\section{METHODS}
\subsection{Theoretical Framework}
\vspace{0.5em}
This was a descriptive observational study utilizing an existing dataset of 193 COVID-19 survivors.
The dataset included a range of demographic, behavioral, and psychological variables collected retrospectively following participants' discharge from the hospital.
The primary objective of the study was to investigate the relationship between psychological factors—depression, anxiety, stress, and optimism—and COVID-19 test outcomes.

\vspace{1em}
\subsection{Research Methodology}
\vspace{0.5em}
Descriptive statistics were calculated to summarize the dataset, providing measures of central tendency such as mean and median, and variability, including standard deviation and interquartile range.
Normality tests, specifically Shapiro-Wilk tests, were performed to evaluate the distribution of the psychological variables. This assessment determined the appropriateness of parametric versus non-parametric statistical analyses.
\\
Correlation analyses were conducted using Spearman's rank correlation coefficient to explore relationships among depression, anxiety, stress, and optimism, particularly due to deviations from normality. Mann-Whitney U tests were used to assess group differences in psychological variables between participants who tested positive for COVID-19 and those who did not. These non-parametric tests were selected due to the nature of the data and its distribution.

\vspace{1em}
\subsection{Statistical Analysis}
\vspace{0.5em}
The analysis of the dataset in relation to the psychological factors—stress, anxiety, depression, and optimism—reveals interesting insights into their potential influence on COVID-19 test status. The Spearman correlation results suggest significant associations between various psychological variables, but the relationship with COVID-19 status is relatively weak, especially when considering the null hypothesis that psychological factors do not influence the likelihood of testing positive or negative for COVID-19.\newline
Starting with the psychological variables, stress is significantly correlated with both anxiety and depression (r = 0.774, p < 0.001 and r = 0.807, p < 0.001, respectively). This indicates that individuals experiencing higher levels of stress are more likely to report higher levels of anxiety and depression. Additionally, optimism shows negative correlations with stress, anxiety, and depression (r = -0.375, r = -0.392, and r = -0.496, respectively), suggesting that individuals with lower levels of optimism tend to experience higher levels of stress, anxiety, and depression. These correlations, however, are not entirely surprising given the established links between these psychological factors in the broader literature.
When considering the influence of these psychological factors on COVID-19 test results, the correlations with the variable Tested positive for Covid are relatively weak. Stress shows a small negative correlation (r = -0.148, p = 0.041), which suggests a very modest inverse relationship between stress levels and testing positive for COVID-19. However, this correlation is far from strong and does not provide evidence to support a significant psychological influence on COVID-19 status. Similarly, anxiety (r = -0.108, p = 0.134) and depression (r = -0.201, p = 0.005) also show weak negative correlations with COVID-19 test results, indicating that higher levels of anxiety and depression may be slightly more common in individuals who test negative, though these relationships are still modest in size. On the other hand, optimism is positively correlated with testing positive for COVID-19 (r = 0.237, p < 0.001), suggesting a small but significant relationship between higher optimism and a higher likelihood of testing positive for the virus. This finding may point to an interesting, albeit small, trend, but it does not imply causation.
Therefore, while significant correlations between psychological variables themselves exist, the relationship between these variables and the likelihood of testing positive for COVID-19 remains weak. The correlation between optimism and testing positive for COVID-19 is the only significant finding, although it is small in magnitude. This analysis suggests that while psychological factors are undeniably interconnected, their influence on COVID-19 test outcomes is minimal, supporting the null hypothesis that stress, anxiety, depression, and optimism do not significantly affect the likelihood of testing positive for the virus.

\end{document}